%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% 13 Jun 2016                                                                 %
% 0_researchnotes.tex                                                         %
% by: David Clarke                                                            %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% Normal preamble stuff.
\documentclass[12pt]{book}
\usepackage[title]{appendix}
\usepackage{mathtools,tensor,physics,slashed,textcomp}

% davitex and required packages
\usepackage{amsmath,amssymb,graphicx,multirow,tabularx}
\usepackage{davitex}

% If you click on citations or equation references, you will be take
% to the corresponding page. 
\usepackage{hyperref}
\hypersetup{
  colorlinks=true,
  linkcolor=black,
  citecolor=magenta,
  urlcolor=blue,
}

% Put references at the end of each chapter, directly after the
% end of the chapter, call them "References" instead of
% "Bibliography", and use numbers with square brackets.
\usepackage{url,natbib,chapterbib}
\let\oldbibliography\bibliography
\renewcommand{\bibliography}[1]{{%
  \let\chapter\section%
  \oldbibliography{#1}}}
\renewcommand\bibname{References}
\setcitestyle{numbers,square}

% Make index.
\usepackage{makeidx}
\makeindex
\usepackage[totoc]{idxlayout}

% Prevent hyphenation.
\setlength\parindent{15pt}
\overfullrule=2cm

% Theorems and stuff.
\usepackage{amsthm}
\usepackage[many]{tcolorbox}
\tcbuselibrary{theorems}
\newtcbtheorem[number within=section]{theorem}{Theorem}{%
  breakable,
  enhanced,
  before skip=10pt,
  colback=green!5,
  colframe=green!35!black,
  fonttitle=\bfseries}{thm}
\newtcbtheorem[number within=section]{proposition}{Proposition}{%
  breakable,
  enhanced,
  before skip=10pt,
  colback=green!5,
  colframe=green!35!black,
  fonttitle=\bfseries}{prp}
\newtcbtheorem[number within=section]{lemma}{Lemma}{%
  breakable,
  enhanced,
  before skip=10pt,
  colback=green!5,
  colframe=green!35!black,
  fonttitle=\bfseries}{lem}
\newtcbtheorem[number within=section]{corollary}{Corollary}{%
  breakable,
  enhanced,
  before skip=10pt,
  colback=green!5,
  colframe=green!35!black,
  fonttitle=\bfseries}{cor}
\newtcbtheorem[number within=section]{example}{Example}{%
  breakable,
  enhanced,
  before skip=10pt,
  colback=blue!5,
  colframe=blue!35!black,
  fonttitle=\bfseries}{cor}
\theoremstyle{definition}
\newtheorem*{defn}{Definition}
%\newtheorem*{example}{Example}

% Some environments for some front matter.
\newenvironment{frontstuff}
  {\centering\chapter*{}}
  {\clearpage}
\newenvironment{listofabbrev}
  {\chapter*{List of Abbreviations}}
  {\clearpage}

% Note that book class by default is formatted to be printed back-to-back.
\title{\bf David's Research Notes} 
\author{David Clarke}

\begin{document}
\frontmatter                            % roman page no., suppress chapter
\pagenumbering{gobble}                  % suppress page no.
\maketitle                              % print title page

\begin{frontstuff} % ----------------------------------------------- FRONT STUFF

\section*{Math Symbols}
\begin{tabular}{ll}
$\exists$       & There exists \\
$\forall$       & For all \\
$\in$           & Is a member of the set \\
$\leq$          & Less than or equal to; is a subgroup of\\
$\approx$       & Approximate equality \\
$\cong$         & Isomorphic\\
$\equiv$        & Is defined as; modular congruency \\
$\sim$          & Equivalence relation \\
$\simeq$        & Asymptotic equality \\
$\mathbb{C}$    & Complex numbers \\
$\co_x$         & $\cos(x)$ \\
$\log$          & Natural logarithm \\
$\mathbb{N}$    & Natural numbers \\
$\mathcal{O}$   & Big O notation; topology on a set \\ 
$\mathbb{Q}$    & Rational numbers \\
$q$             & Gaussian or Student difference test; goodness-of-fit \\
$\mathbb{R}$    & Real numbers \\
$\s_x$          & $\sin(x)$ \\
$\SU(N)$        & Special unitary group of degree $N$ \\
$S_V$           & Symmetric group on the set $V$ \\
$\tauint$       & Integrated autocorrelation time \\
$\mathbb{Z}$    & Integers \\
$\mathbb{Z}_n$  & Integers modulo $n$ \\
\end{tabular}
\clearpage

\section*{Physics Symbols}
\begin{tabular}{ll}
$a$             & Lattice spacing \\
$\beta$         & $\SU(N_c)$ coupling constant; beta function \\
$\chi$          & Topological susceptibility; Polyakov loop susceptibility\\
$c$             & Speed of light\\
$G$             & A group; Newton's gravitational constant\\
$g$             & A group element; $\SU(N_c)$ bare coupling\\
$\hbar$         & Planck's constant \\
$k_B$           & Boltzmann's constant \\
$N_c$           & Number of colors \\
$N_f$           & Number of fermion flavors \\
$N_s$           & Lattice extension in a spatial direction \\
$N_\tau$        & Lattice extension in temperature/Euclidean time direction\\
$\sigma$        & String tension\\
$\sigma_i$      & A Pauli matrix\\
$T_c$           & Deconfining phase transition temperature\\
\end{tabular}
\clearpage

\section*{High Energy Glossary}
\begin{tabularx}{\linewidth}{lL}
Baryogenesis&
  Physical processes that must have occurred
  during the early universe that are responsible for the imbalance between
  the amount of matter and antimatter.
\vspace{2mm}\\
Baryon& 
  Three quark hadron.
\vspace{2mm}\\
Baryon number (B)& 
  (\# baryons $-$ \# anti-baryons)
\vspace{2mm}\\
Bottomness (B$'$)& 
  $-$(\# bottom quarks $-$ \# bottom anti-quarks)
\vspace{2mm}\\
Branching ratio& 
  Ratio of decay rate of particle along a particular channel
  to the total decay rate of that particle.
\vspace{2mm}\\
Charm (C)& 
  (\# charm quarks $-$ \# charm anti-quarks)
\vspace{2mm}\\
Effective field theory&
  Field theory that is valid only at certain energy scales. The SM is 
  therefore an effective field theory.
\vspace{2mm}\\
Field equation&
  Any partial differential equation describing the dynamics of
  some physical field.
\vspace{2mm}\\
Hadron& 
  Color neutral particle made of multiple quarks. See e.g. meson, baryon, 
  nucleon.
\vspace{2mm}\\
Lattice units&
  Units where $a=1$ and is dimensionless.
\vspace{2mm}\\
Meson& 
  Two quark hadron.
\vspace{2mm}\\
Naturalness&
  The property that free parameters or physical constants appearing in a theory
  should take relative values of order 1.
\vspace{2mm}\\
Nucleon& 
  Baryons found in atomic nuclei.
\vspace{2mm}\\
Physical units& 
  Units where $a$ carries length. 
\vspace{2mm}\\
Planck mass&
  Mass at which the gravitational coupling becomes of order one.
  $$
    \alpha_G=\frac{GM_P^2}{\hbar c}\sim1~\Rightarrow~
     M_P\sim\sqrt{\frac{\hbar c}{G}}\approx10^{19}~\text{GeV}
  $$
\end{tabularx}
\clearpage
\begin{tabularx}{\linewidth}{lL}
Semi-leptonic decay& 
  Weak decay of a hadron producing a lepton, corresponding neutrino,
  and one or more hadrons. 
\vspace{2mm}\\
Strangeness (S)& 
  $-$(\# strange quarks $-$ \# strange anti-quarks)
\vspace{2mm}\\
Topness (T)& 
  (\# top quarks $-$ \# top anti-quarks)
\end{tabularx}
\clearpage

\section*{Some Important Particles}
\begin{tabularx}{\linewidth}{rLlc}
{\it Name} & {\it Significance} & {\it Symbol} & {\it Composition}
\vspace{2mm}\\
B meson: & Light mesons with bottomness 1.
                                     & $B^0$ &   $d\bar{b}$, 
                                                   $b\bar{d}$ \\
         & Can use weak decays to extract
                                     & $B^+$ &   $u\bar{b}$ \\
         & some CKM matrix elements. & $B^-$ &   $b\bar{u}$
\vspace{2mm}\\
D meson: & Light mesons with charm 1. Can
                                     & $D^0$ &   $c\bar{u}$, 
                                                   $u\bar{c}$ \\
         & use weak decays to extract some
                                     & $D^+$ &   $c\bar{d}$ \\
         & CKM matrix elements.      & $D^-$ &   $d\bar{c}$
\vspace{2mm}\\
Eta:     & $\eta-\eta'$ mass difference can be
                                     & $\eta$  &   $u\bar{u}
                                                   +d\bar{d}
                                                  -2s\bar{s}$ \\
         & explained by topology.    & $\eta'$ &   $u\bar{u}
                                                   +d\bar{d}
                                                   +s\bar{s}$
\vspace{2mm}\\
Kaon:    & Light mesons with strangeness 1.
                                     & $K^0$ &   $d\bar{s}$, 
                                                   $s\bar{d}$ \\
         & Can use weak decays to extract 
                                     & $K^+$ &   $u\bar{s}$ \\
         & some CKM matrix elements. Its 
                                     & $K^-$ &   $s\bar{u}$ \\
         & decay constant is used as a 
                                     &       &              \\
         & reference in lattice calculations.
                                     &       &              
\vspace{2mm}\\
Pion:    & Lightest mesons. ``Would-be"
                                     & $\pi^0$ &   $u\bar{u}$,
                                                   $d\bar{d}$ \\
         & Goldstone bosons of chiral& $\pi^+$ &   $u\bar{d}$ \\
         & symmetry breaking.        & $\pi^-$ &   $d\bar{u}$
\vspace{2mm}\\
Psion:   & Discovery of the charm quark.
         & $\text{J}/\psi$                     &   $c\bar{c}$ 
\vspace{2mm}\\
Upsilon: & Discovery of the bottom quark.
         & $\Upsilon$                          &   $b\bar{b}$ \\
\end{tabularx}
\clearpage

\end{frontstuff} % ----------------------------------------------------- PREFACE

\chapter{Preface}
This ``book" is a collection of research notes that I made while researching in
physics. The first chapters are dedicated to mainly mathematics, and the last
chapters are dedicated to mainly physics, with an aim toward lattice field
theory. Attached at the end are also some appendices. 
Appendix~\ref{ap:spec_math} includes some special topics in mathematics 
that, while sometimes useful for a physicist, don't really fit in one of 
the mathematics chapters at the beginning. Appendix~\ref{ap:spec_phys} includes
special topics in physics that I find interesting, but aren't directly
relevant to any research I work on. As I continue researching, I may want
to develop some of these appendices into full chapters, if they ever
become directly relevant to my research.

Throughout this book, I assume the reader has some basic knowledge of 
mathematics, for instance what it means for operations to be $associative$ or 
$commutative$. The reader will also find it useful to have some basic knowledge
of calculus, linear algebra, differential equations, and complex analysis.
(Just enough to get by in undergraduate level physics classes.) 

\tableofcontents                        % print table of contents
\mainmatter                             % arabic page no., include chapter

% These are the chapters of the book.
\include{ch_group_theory}
%\include{ch_differential_geometry}
\include{ch_probability_and_statistics}
\include{ch_statistical_physics}
%\include{ch_QFT}
\include{ch_LGT_prelim}
\include{ch_LGT_MCMC}
\include{ch_LFT_fermions}

% Here are the appendices.
\begin{appendices}
\include{ap_spec_math}
\include{ap_spec_phys}
\include{ap_spec_neutrino}
\include{ap_spec_higgs}
\end{appendices}

\printindex

\end{document}
